---
title: "Thesis"
author: "Jason Leff"
date: "2025-09-19"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(scatterplot3d)
library(MASS)
library(caret)
library(glmnet)
library(car)
library(MLmetrics)
library(nnet)
library(pROC)
library(multiROC)
library(cluster)
library(rms)
library(leaps)
library(e1071)
library(randomForest)
library(xgboost)
library(factoextra)
library(nnet)
library(glmnetcr)
library(ordinalNet)
library(pracma)
library(pheatmap)
library(rpart)
library(rpart.plot)
```

# Setup
```{r}
rm(list=ls())
data <- read.csv('data_with_pre_data.csv')
set.seed(67)
```
## Initializing Functions
```{r}
heatmap <- function(ConfMatrix, color, title) { #Unused heatmap function
  cm_table <- as.data.frame(ConfMatrix$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = color) +
  theme_minimal() +
  labs(title = paste0("Confusion Matrix Heatmap (", title, ")"),
       x = "Actual Class", y = "Predicted Class")
}
```

```{r}
within_one_class_pct <- function(ConfusionMatrix) { #Unused ordinal statistic calculator
  classes <- rownames(ConfusionMatrix)
  n_class <- length(classes)
  idx <- seq_len(n_class)
  total <- sum(ConfusionMatrix)
  count_valid <- 0
  for (i in idx) {
    valid_pred <- i
    if (i == 1) {valid_pred <- c(valid_pred, 2)} 
    else if (i == n_class) {valid_pred <- c(valid_pred, n_class - 1)} 
    else {valid_pred <- c(valid_pred, i - 1, i + 1)}
    count_valid <- count_valid + sum(ConfusionMatrix[i, valid_pred])
  }
  count_valid / total
}
```

```{r}
macro_auc <- function(y_true, prob_matrix) { #Used to calculate AUC for each model
  classes <- levels(y_true)
  aucs <- numeric(length(classes))
  for (i in seq_along(classes)) {
    cls <- classes[i]
    y_binary <- as.numeric(y_true == cls)
    p <- prob_matrix[, cls]

    roc_obj <- try(roc(y_binary, p, quiet = TRUE), silent = TRUE)

    if (inherits(roc_obj, "try-error") || is.null(roc_obj$auc)) {
      aucs[i] <- NA_real_
    } else {
      aucs[i] <- as.numeric(roc_obj$auc)
    }
  }
  mean(aucs, na.rm = TRUE)
}
```

## Normalize to Per Minute Data
```{r}
normalize_stats <- function(df) {
  nba_start <- which(names(df) == "NBA_MIN") #split the data
  nba_end   <- which(names(df) == "PRE_G")
  
  for (i in (nba_start+1):(nba_end-1)) {
    colname <- names(df)[i]
    if (!grepl("PCT", colname, ignore.case = TRUE) && is.numeric(df[[colname]])) { #divide total columns by minute column
      df[[colname]] <- round(as.numeric(df[[colname]]) / as.numeric(df[["NBA_MIN"]]),3)
    }
  }
  pre_start <- which(names(df) == "PRE_MP")
  for (i in (pre_start+1):ncol(df)) { #repeat for college
    colname <- names(df)[i]
    if (!grepl("PCT", colname, ignore.case = TRUE) && is.numeric(df[[colname]])) {
      df[[colname]] <- round(as.numeric(df[[colname]]) / as.numeric(df[["PRE_MP"]]),3)
    }
  }
  return(df)
}
per_min_data <- normalize_stats(data)
```

## PRE-NBA FT% and 3P% vs NBA 3P% filtered for 25 attempts
```{r}
data_filtered2 <- data %>%
  filter(PRE_3PA >= 25 & PRE_FTA >= 25 & NBA_FG3A >= 25)

clean_data <- na.omit(data_filtered2[c("PLAYER","NBA_FG3_PCT", "PRE_3P_PCT", "PRE_FT_PCT", "PRE_COL_or_INT")])
model <- lm(NBA_FG3_PCT ~ PRE_3P_PCT + PRE_FT_PCT, data = clean_data)
summary(model)

color_factor <- as.factor(clean_data$PRE_COL_or_INT)

my_colors <- c(
  "COL" = "#1f77b4",   # blue
  "INT" = "#d62728"    # red
)

point_colors <- my_colors[color_factor]
s3d <- scatterplot3d(
  clean_data$PRE_3P_PCT,
  clean_data$PRE_FT_PCT,
  clean_data$NBA_FG3_PCT,
  color = point_colors,
  pch = 16,
  xlab = "Pre-3P%, >= 25 Attempts",
  ylab = "Pre-FT%, >= 25 Attempts",
  zlab = "NBA 3P%, >= 25 Attempts",
  main = "Pre-NBA 3P% and FT% vs. NBA 3P%, minimum 25 attempts"
)

s3d$plane3d(model, lty.box = "solid")
legend(
  "topright",
  legend = levels(color_factor),
  col = my_colors,
  pch = 16
)

```

# Clustering
## Clustering to Create Classes
```{r}
indices <- which(data$NBA_FG3A >= 10 & data$NBA_FG3_PCT > 0.05) #Use a bit less strict filtering
clust_data <- na.omit(per_min_data[indices, c("NBA_FG3A", "NBA_FG3_PCT")])
clust_data_scaled <- scale(clust_data)

sil_width <- numeric(8)
for (k in 4:8) { #use silhouette method
  set.seed(123)
  km <- kmeans(clust_data_scaled, centers = k, nstart = 25)
  ss <- silhouette(km$cluster, dist(clust_data_scaled))
  sil_width[k] <- mean(ss[, 3])
}

plot(4:8, sil_width[4:8], type = "b", pch = 19, #plot results
     xlab = "Number of clusters K",
     ylab = "Average silhouette width",
     main = "Silhouette Method for Optimal K")

best_k <- which.max(sil_width) #set best k

set.seed(123)
km_best <- kmeans(clust_data_scaled, centers = best_k, nstart = 25) #graph the best k clusters

clust_plot <- data.frame(clust_data, cluster = factor(km_best$cluster))

ggplot(clust_plot, aes(x = NBA_FG3A, y = NBA_FG3_PCT, color = cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = paste("K-means Clustering with K =", best_k,
                     " (Silhouette Method)")) +
  theme_minimal()
```

## Individual Clustering
```{r}
y <- per_min_data$NBA_FG3A[indices]
sel <- indices[ which(!is.na(per_min_data$NBA_FG3A[indices])) ]
x <- na.omit(per_min_data$NBA_FG3A[indices])
x_scaled <- scale(x)
d <- dist(x_scaled)
hc <- hclust(d, method = "ward.D2")

# Plot dendrogram with cutoff lines and rectangles for k = 3 to 8
plot(hc, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (NBA_FG3A)\nwith Cluster Cutoffs (k = 3 to 5)")

# Get the height range of the dendrogram
height_range <- max(hc$height)

# Add horizontal cutoff lines and colored rectangles for each k
colors <- c("red", "blue", "green3")

for (i in 1:length(3:5)) {
  k <- (3:5)[i]
  
  # Find the height at which cutting produces exactly k clusters
  # This is the midpoint between merge heights for k and k-1 clusters
  if(k < length(hc$height)) {
    cutoff_height <- (hc$height[length(hc$height) - k + 1] + 
                      hc$height[length(hc$height) - k + 2]) / 2
  } else {
    cutoff_height <- hc$height[1] + 0.1 * diff(range(hc$height))
  }
  
  # Add horizontal cutoff line
  abline(h = cutoff_height, col = colors[i], lty = 2, lwd = 1.5)
  
  # Add label for the cutoff
  text(x = length(hc$order) * 0.98, y = cutoff_height, 
       labels = paste0("k = ", k), 
       col = colors[i], cex = 0.8, adj = c(1, -0.5))
  
  # Add colored rectangles for the clusters
  rect.hclust(hc, k = k, border = colors[i])
}

# Add a legend
legend("topright", legend = paste("k =", 3:5), 
       col = colors, lty = 2, lwd = 1.5, cex = 0.8)

# Compute silhouette widths
sil_width <- numeric(20)

for (k in 3:8) {
  clusters <- cutree(hc, k)
  ss <- silhouette(clusters, d)
  sil_width[k] <- mean(ss[, 3])
}

# Plot silhouette results
plot(3:8, sil_width[3:8], type = "b", pch = 19, cex = 1.5,
     xlab = "Number of clusters k",
     ylab = "Average silhouette width",
     main = "Silhouette Method (NBA_FG3A)",
     xaxt = "n")  # Suppress default x-axis to customize
axis(1, at = 3:8)  # Add custom x-axis with integer ticks
```
```{r}
best_k <- which.max(sil_width)
clusters_best <- cutree(hc, k = best_k)

plot(
  x,
  jitter(rep(0, length(x)), amount = 0.05),
  col = clusters_best, pch = 19, xlab = "NBA_FG3A", ylab = "",
  yaxt = "n", main = paste("Hierarchical Clustering on NBA_FG3A (k =", best_k, ")")
)

```

```{r}
cluster_medians <- tapply(x, clusters_best, median)
ordered_clusters <- names(sort(cluster_medians, decreasing = FALSE))
new_label_map <- setNames(seq_along(ordered_clusters), ordered_clusters)
clusters_ordered <- as.integer(new_label_map[as.character(clusters_best)])
per_min_data$Class_3PA <- NA_integer_
per_min_data$Class_3PA[sel] <- clusters_ordered
per_min_data$Class_3PA[is.na(per_min_data$Class_3PA)] <- 1
per_min_data$Class_3PA <- factor(per_min_data$Class_3PA,
                                             levels = as.character(1:best_k),
                                             ordered = TRUE)
round(tapply(x, per_min_data$Class_3PA[sel], mean),3)
table(per_min_data$Class_3PA, useNA = "ifany")
#head(per_min_data)
```

```{r}
y <- per_min_data$NBA_FG3_PCT[indices]
not_na_idx <- which(!is.na(y))
sel <- indices[ which(!is.na(per_min_data$NBA_FG3_PCT[indices])) ]
x <- na.omit(data$NBA_FG3_PCT[indices])
x_scaled <- scale(x)
d <- dist(x_scaled)
hc <- hclust(d, method = "ward.D2")
plot(hc, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (NBA_FG3_PCT)")
sil_width <- numeric(20)

for (k in 4:8) {
  clusters <- cutree(hc, k)
  ss <- silhouette(clusters, d)
  sil_width[k] <- mean(ss[, 3])
}

plot(4:8, sil_width[4:8], type = "b", pch = 19,
     xlab = "Number of clusters k",
     ylab = "Average silhouette width",
     main = "Silhouette Method (NBA_FG3_PCT)")
```

```{r}
best_k <- which.max(sil_width)
clusters_best <- cutree(hc, k = best_k)

plot(x, rep(0, length(x)), col = clusters_best, pch = 19,
     xlab = "NBA_3P%", ylab = "",
     yaxt = "n", main = paste("Hierarchical Clustering on NBA_FG3_PCT (k =", best_k, ")"))
```
```{r}
cluster_medians <- tapply(x, clusters_best, median)
ordered_clusters <- names(sort(cluster_medians, decreasing = FALSE))
new_label_map <- setNames(seq_along(ordered_clusters), ordered_clusters)
clusters_ordered <- as.integer(new_label_map[as.character(clusters_best)])
per_min_data$Class_3P_PCT <- NA_integer_
per_min_data$Class_3P_PCT[sel] <- clusters_ordered
per_min_data$Class_3P_PCT[is.na(per_min_data$Class_3P_PCT)] <- 1
per_min_data$Class_3P_PCT <- factor(per_min_data$Class_3P_PCT,
                                             levels = as.character(1:best_k),
                                             ordered = TRUE)
round(tapply(x, per_min_data$Class_3P_PCT[sel], median),3) 
table(per_min_data$Class_3P_PCT, useNA = "ifany")
#head(per_min_data)
```

## Assign Clusters to the dataset
```{r}
# Assign the cluster numbers
per_min_data$Cluster3Shooting <- NA_integer_ 
valid_rows <- indices[complete.cases(per_min_data[indices, c("NBA_FG3A", "NBA_FG3_PCT")])]
per_min_data$Cluster3Shooting[valid_rows] <- km_best$cluster
table(per_min_data$Cluster3Shooting)
```
```{r}
# Rearrange for any functions that may still use ordinal
per_min_data$Cluster3Shooting_reordered <- dplyr::case_when(
  per_min_data$Cluster3Shooting == 2 ~ 1L,
  per_min_data$Cluster3Shooting == 5 ~ 2L,
  per_min_data$Cluster3Shooting == 4 ~ 3L,
  per_min_data$Cluster3Shooting == 1 ~ 4L,
  per_min_data$Cluster3Shooting == 3 ~ 5L,
  TRUE ~ 1L 
)

per_min_data$Cluster3Shooting_reordered <- dplyr::case_when(
  per_min_data$NBA_FG3_PCT < 0.3 & per_min_data$Cluster3Shooting_reordered == 5L ~ 2L,
  TRUE ~ per_min_data$Cluster3Shooting_reordered
)

```
```{r}
# Check summary
table(per_min_data$Cluster3Shooting_reordered)
```

```{r}
# Set labels
class_levels <- c("LowLow", "MidLow", "LowMid", "MidHigh", "HighHigh")
per_min_data$Cluster3Shooting_reordered <- factor(per_min_data$Cluster3Shooting_reordered,
                                        levels = 1:5,
                                        labels = class_levels,
                                        ordered = TRUE)
```

```{r}
# Graph clusters with better colors
ggplot(per_min_data, aes(x = NBA_FG3A, y = NBA_FG3_PCT, color = Cluster3Shooting_reordered)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(
    values = c("LowLow" = "red",
               "MidLow" = "orange",
               "LowMid" = "yellow",
               "MidHigh" = "green",
               "HighHigh" = "darkgreen")
  ) +
  labs(
    title = "NBA 3PA Per Minute vs 3P% Colored by Combined Class",
    x = "3PA Per Minute",
    y = "3P Percentage"
  ) +
  theme_minimal()
```

```{r}
ggplot(per_min_data, aes(x = NBA_FG3A, y = NBA_FG3_PCT)) +
  geom_point(aes(color = factor(Class_3P_PCT), size = factor(Class_3PA)), alpha = 0.7) +
  scale_color_manual(name = "3P% Class", values = c("1" = "darkred", "2" = "red", "3" = "orange",
      "4" = "yellow", "5" = "lightgreen", "6" = "darkgreen"),
    labels = c("1 (Low)", "2", "3", "4", "5", "6 (High)")) +
  scale_size_manual(name = "3PA Class", values = c("1" = 1, "2" = 2, "3" = 3, "4" = 4,
      "5" = 5),
    labels = c("1 (Low)", "2", "3", "4", "5 (High)")) +
  labs(title = "NBA 3PA Per Minute vs NBA 3 Point Percentage", x = "3PA Per Minute", y = "3P Percentage") +
  theme_minimal()
```


# Prediction on both 3PA and 3P% Class
## Initialize Train / Test Split
```{r}
# Specific set of predictors mentioned in report
X <- per_min_data[, c(8,13,38:40,42,43,45,46,48:50, 52:55, 56:66)]
X[is.na(X)] <- 0
y <- factor(per_min_data[, ncol(per_min_data)], levels = class_levels)

df <- data.frame(X, response = y)
df_scaled <- data.frame(scale(df[, -ncol(df)]), response = df$response)

# Split train and test
set.seed(67)
train_index <- createDataPartition(df_scaled$response, p = 0.8, list = FALSE)
train_data <- df_scaled[train_index, ]
test_data  <- df_scaled[-train_index, ]
```



## Backwards Stepwise Selection
```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
full_model <- orm(response ~ ., data = train_data)
(step_model <- fastbw(fit = full_model, rule="p", type="individual", sls=0.001))
```
## Check for collinearity
```{r}
vif(step_model)
```
```{r}
predictors <- df_scaled[, c("PRE_FGA", "PRE_FG_PCT", "PRE_3PA", 
    "PRE_FTA", "PRE_DRB", "PRE_BLK", "PRE_TOV", "PRE_PTS")]
cor_matrix <- cor(predictors)
print(round(cor_matrix, 3))
```
Our main issue is between PPG and FGA, check both.
```{r}
noPTS <- polr(response ~ PRE_FGA+PRE_FG_PCT+PRE_3PA+PRE_FTA+PRE_DRB+PRE_BLK+PRE_TOV, data = train_data, Hess = TRUE)
noFGA <- polr(response ~ PRE_FG_PCT+PRE_3PA+PRE_FTA+PRE_DRB+PRE_BLK+PRE_TOV+PRE_PTS, data = train_data, Hess = TRUE)

AIC(noPTS);AIC(noFGA)
```
Because the no FGA model has the lower AIC, it is the better model because we chose AIC as our criterion.
```{r}
summary(noFGA)
vif(noFGA)
predictors <- df_scaled[, c("PRE_FG_PCT", "PRE_3PA", 
    "PRE_FTA", "PRE_DRB", "PRE_BLK", "PRE_TOV", "PRE_PTS")]
cor_matrix <- cor(predictors)
print(round(cor_matrix, 3))
```

## Initializing Storing Results
```{r}
model_results <- data.frame(
  Model = character(),
  Accuracy = numeric(),
  Precision = numeric(),
  Recall = numeric(),
  F1 = numeric(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)
```

## LDA 
```{r}
set.seed(22222)
# Run cross validation
train_control <- trainControl(method = "cv", number = 10,
                              summaryFunction = multiClassSummary)

lda_cv <- train(response ~ PRE_FG_PCT + PRE_3PA + PRE_FTA + PRE_DRB + 
                           PRE_BLK + PRE_TOV + PRE_PTS, data = train_data,
                method = "lda",
                trControl = train_control)

print(lda_cv)
cat("Cross-validated Accuracy (training set):", max(lda_cv$results$Accuracy), "\n")

# Set fit model
lda_fit <- lda(response ~ PRE_FG_PCT + PRE_3PA + PRE_FTA + PRE_DRB + 
                           PRE_BLK + PRE_TOV + PRE_PTS,
               data = train_data)

pred_obj <- predict(lda_fit, test_data)

all_pred   <- factor(pred_obj$class, levels = class_levels)
all_actual <- factor(test_data$response, levels = class_levels)
all_probs  <- pred_obj$posterior

# Find CM
conf_mat_lda <- confusionMatrix(all_pred, all_actual, mode = "prec_recall")
print(conf_mat_lda)

# Calculate Accuracy and AUC
LDA_acc <- mean(as.character(all_pred) == as.character(all_actual))
cat("Test Accuracy (LDA):", LDA_acc, "\n")

overall_multiclass <- multiclass.roc(all_actual, all_probs)
cat("Overall multiclass AUC (test set):", overall_multiclass$auc, "\n")

# Graph AUC / Calculate Macro AUC
fpr_grid <- seq(0, 1, length.out = 200)
tpr_matrix <- matrix(NA, nrow = length(fpr_grid), ncol = length(class_levels))

for (i in seq_along(class_levels)) {
  class <- class_levels[i]
  binary_actual <- factor(ifelse(all_actual == class, class, "Other"),
                          levels = c("Other", class))
  class_probs <- all_probs[, class]
  
  roc_obj <- roc(response = binary_actual,
                 predictor = class_probs,
                 levels = c("Other", class),
                 positive = class,
                 quiet = TRUE)
  
  cat("AUC of", class, ":", auc(roc_obj), "\n")
  
  fpr <- 1 - roc_obj$specificities
  tpr <- roc_obj$sensitivities
  
  interp_tpr <- approx(fpr, tpr, xout = fpr_grid, ties = mean, rule = 2)$y
  tpr_matrix[, i] <- interp_tpr
}
macro_tpr_lda <- rowMeans(tpr_matrix, na.rm = TRUE)
macro_auc_lda <- trapz(fpr_grid, macro_tpr_lda)
cat("Macro AUC (LDA test set):", macro_auc_lda, "\n")
```

```{r}
model_results <- rbind(model_results, data.frame(
  Model = "LDA",
  Accuracy = LDA_acc,
  Precision = mean(conf_mat_lda$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(conf_mat_lda$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(conf_mat_lda$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc_lda
))
```

```{r}
cm_table <- as.data.frame(conf_mat_lda$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "darkorange") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap (LDA)",
       x = "Actual Class", y = "Predicted Class")
```
```{r}
# LDA Decision Boundary Visualization
# Project data onto first two LDA dimensions
lda_proj_train <- as.data.frame(predict(lda_fit, train_data)$x[, 1:2])
lda_proj_train$class <- train_data$response
colnames(lda_proj_train)[1:2] <- c("LD1", "LD2")

# Create grid for decision regions
ld1_range <- seq(min(lda_proj_train$LD1) - 1, max(lda_proj_train$LD1) + 1, length.out = 150)
ld2_range <- seq(min(lda_proj_train$LD2) - 1, max(lda_proj_train$LD2) + 1, length.out = 150)
grid <- expand.grid(LD1 = ld1_range, LD2 = ld2_range)

# Get training means for each predictor
train_means <- colMeans(train_data[, c("PRE_FG_PCT", "PRE_3PA", "PRE_FTA", "PRE_DRB", 
                                        "PRE_BLK", "PRE_TOV", "PRE_PTS")])

# Reconstruct original variables from LD scores
grid_orig <- matrix(NA, nrow = nrow(grid), ncol = 7)
colnames(grid_orig) <- names(train_means)

for(i in 1:nrow(grid)) {
  grid_orig[i, ] <- train_means + (grid$LD1[i] * lda_fit$scaling[, 1] + 
                                     grid$LD2[i] * lda_fit$scaling[, 2])
}

grid_pred <- predict(lda_fit, as.data.frame(grid_orig))$class
grid$class <- factor(grid_pred, levels = class_levels)

# Plot
ggplot() +
  geom_tile(data = grid, aes(x = LD1, y = LD2, fill = class), alpha = 0.3) +
  geom_point(data = lda_proj_train, aes(x = LD1, y = LD2, color = class), size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00"),
                    drop = FALSE) +
  scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00"),
                     drop = FALSE) +
  labs(title = "LDA Decision Regions for Combined Class",
       subtitle = paste0("Cross-validated Accuracy: ", round(max(lda_cv$results$Accuracy), 3)),
       x = "Linear Discriminant 1",
       y = "Linear Discriminant 2",
       fill = "Class",
       color = "Class") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11))
```


## QDA
```{r}
# Same steps as LDA
set.seed(22222)
train_control <- trainControl(method = "cv", number = 10,
                              summaryFunction = multiClassSummary)

qda_cv <- train(response ~ PRE_FG_PCT + PRE_3PA + PRE_FTA + PRE_DRB + 
                           PRE_BLK + PRE_TOV + PRE_PTS,
                data = train_data,
                method = "qda",
                trControl = train_control)

print(qda_cv)
cat("Cross-validated Accuracy (training set):", max(qda_cv$results$Accuracy), "\n")

qda_fit <- qda(response ~ PRE_FG_PCT + PRE_3PA + PRE_FTA + PRE_DRB + 
                           PRE_BLK + PRE_TOV + PRE_PTS,
               data = train_data)


pred_obj <- predict(qda_fit, test_data)

all_pred   <- factor(pred_obj$class, levels = class_levels)
all_actual <- factor(test_data$response, levels = class_levels)
all_probs  <- pred_obj$posterior

conf_mat_qda <- confusionMatrix(all_pred, all_actual, mode = "prec_recall")
print(conf_mat_qda)

QDA_acc <- mean(as.character(all_pred) == as.character(all_actual))
cat("Test Accuracy (QDA):", QDA_acc, "\n")

overall_multiclass <- multiclass.roc(all_actual, all_probs)
cat("Overall multiclass AUC (QDA test set):", overall_multiclass$auc, "\n")

fpr_grid <- seq(0, 1, length.out = 200)
tpr_matrix <- matrix(NA, nrow = length(fpr_grid), ncol = length(class_levels))

for (i in seq_along(class_levels)) {
  class <- class_levels[i]
  binary_actual <- factor(ifelse(all_actual == class, class, "Other"),
                          levels = c("Other", class))
  class_probs <- all_probs[, class]
  
  roc_obj <- roc(response = binary_actual,
                 predictor = class_probs,
                 levels = c("Other", class),
                 positive = class,
                 quiet = TRUE)
  
  cat("AUC of", class, ":", auc(roc_obj), "\n")
  
  fpr <- 1 - roc_obj$specificities
  tpr <- roc_obj$sensitivities
  
  interp_tpr <- approx(fpr, tpr, xout = fpr_grid, ties = mean, rule = 2)$y
  tpr_matrix[, i] <- interp_tpr
}

macro_tpr_qda <- rowMeans(tpr_matrix, na.rm = TRUE)
macro_auc_qda <- trapz(fpr_grid, macro_tpr_qda)
cat("Macro AUC (QDA test set):", macro_auc_qda, "\n")
```


```{r}
model_results <- rbind(model_results, data.frame(
  Model = "QDA",
  Accuracy = QDA_acc,
  Precision = mean(conf_mat_qda$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(conf_mat_qda$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(conf_mat_qda$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc_qda
))
```

```{r}
cm_table <- as.data.frame(conf_mat_qda$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "skyblue") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap (QDA) for Combined Class",
       x = "Actual Class", y = "Predicted Class")
```

```{r}
# QDA Decision Boundary Visualization
# Project data using LDA dimensions for visualization purposes
qda_proj_train <- as.data.frame(as.matrix(train_data[, c("PRE_FG_PCT", "PRE_3PA", "PRE_FTA", "PRE_DRB", 
                                                          "PRE_BLK", "PRE_TOV", "PRE_PTS")]) %*% 
                                  lda_fit$scaling[, 1:2])
qda_proj_train$class <- train_data$response
colnames(qda_proj_train)[1:2] <- c("LD1", "LD2")

# Create grid for decision regions
ld1_range <- seq(min(qda_proj_train$LD1) - 1, max(qda_proj_train$LD1) + 1, length.out = 150)
ld2_range <- seq(min(qda_proj_train$LD2) - 1, max(qda_proj_train$LD2) + 1, length.out = 150)
grid <- expand.grid(LD1 = ld1_range, LD2 = ld2_range)

# Get training means for each predictor
train_means <- colMeans(train_data[, c("PRE_FG_PCT", "PRE_3PA", "PRE_FTA", "PRE_DRB", 
                                        "PRE_BLK", "PRE_TOV", "PRE_PTS")])

# Reconstruct original variables from LD scores
grid_orig <- matrix(NA, nrow = nrow(grid), ncol = 7)
colnames(grid_orig) <- names(train_means)

for(i in 1:nrow(grid)) {
  grid_orig[i, ] <- train_means + (grid$LD1[i] * lda_fit$scaling[, 1] + 
                                     grid$LD2[i] * lda_fit$scaling[, 2])
}

grid_pred <- predict(qda_fit, as.data.frame(grid_orig))$class
grid$class <- factor(grid_pred, levels = class_levels)

# Plot
ggplot() +
  geom_tile(data = grid, aes(x = LD1, y = LD2, fill = class), alpha = 0.3) +
  geom_point(data = qda_proj_train, aes(x = LD1, y = LD2, color = class), size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00"),
                    drop = FALSE) +
  scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00"),
                     drop = FALSE) +
  labs(title = "QDA Decision Regions for Combined Class",
       subtitle = paste0("Cross-validated Accuracy: ", round(max(qda_cv$results$Accuracy), 3)),
       x = "Discriminant 1",
       y = "Discriminant 2",
       fill = "Class",
       color = "Class") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11))
```


## Comparison of LDA vs QDA
```{r}
LDA_acc - QDA_acc
```
```{r}
# Graph AUC curves
plot(fpr_grid, macro_tpr_lda, type = "l", lwd = 2, col = "black",
     xlab = "1 - Specificity (FPR)", ylab = "Sensitivity (TPR)",
     main = paste("Macro-average ROC Curves of LDA and QDA"))
abline(a = 0, b = 1, lty = 2, col = "gray")
lines(fpr_grid, macro_tpr_qda, lwd = 2, col = "red")
legend("bottomright",
       legend = c(paste("LDA (AUC =", round(macro_auc_lda, 3), ")"),
                  paste("QDA (AUC =", round(macro_auc_qda, 3), ")")),
       col = c("black", "red"), lwd = 2)
```

## LASSO with CV for best lambda
```{r}
set.seed(67)
#initialize data in matrices
x_train <- as.matrix(train_data[, -ncol(train_data)]) 
y_train <- train_data$response
x_test  <- as.matrix(test_data[, -ncol(test_data)])
y_test  <- test_data$response

#run CV model
cv_lasso <- cv.glmnet(x_train, y_train, 
                      family = "multinomial",
                      alpha = 1,
                      nfolds = 10)
plot(cv_lasso)

# Check optimal values of lambda
cat("Optimal lambda (min):", cv_lasso$lambda.min, "\n") 
cat("Optimal lambda (1se):", cv_lasso$lambda.1se, "\n")

log_lambda_min <- log(cv_lasso$lambda.min)
log_lambda_1se <- log(cv_lasso$lambda.1se)

# Get the y-axis limits to position the labels
y_range <- range(cv_lasso$cvm + cv_lasso$cvsd, cv_lasso$cvm - cv_lasso$cvsd, na.rm = TRUE)

# Add labels near the top of the plot
text(x = log_lambda_min, 
     y = y_range[2] * 0.95, 
     labels = paste0("Min: ", round(cv_lasso$lambda.min, 4)),
     pos = 2,
     cex = 0.8,
     col = "red")

text(x = log_lambda_1se, 
     y = y_range[2] * 0.88, 
     labels = paste0("1SE: ", round(cv_lasso$lambda.1se, 4)),
     pos = 2,
     cex = 0.8,
     col = "blue")
```

```{r}
# Extract coefficients for lambda.min
coefs <- coef(cv_lasso, s = "lambda.min")

# Combine into a matrix (predictors x classes)
# Remove intercept row and convert to matrix
coef_matrix <- sapply(coefs, function(x) as.matrix(x)[-1, , drop = FALSE])
colnames(coef_matrix) <- levels(as.factor(y_train))
rownames(coef_matrix) <- rownames(coefs[[1]])[-1]  # predictor names without intercept

# Keep only non-zero coefficients (or all if you prefer)
non_zero_rows <- apply(coef_matrix, 1, function(x) any(x != 0))
coef_nz <- coef_matrix[non_zero_rows, , drop = FALSE]

# Plot heatmap
pheatmap(coef_nz, 
         cluster_rows = TRUE, 
         cluster_cols = FALSE, 
         main = paste0("Non-zero Coefficients at lambda_min = ", round(cv_lasso$lambda.min, 4)),
         color = colorRampPalette(c("blue", "white", "red"))(100),
         fontsize_row = 8,
         display_numbers = FALSE)
```

```{r}
class_names <- levels(as.factor(y_train))
par(mfrow = c(2, 3))  # 2x3 grid for 5 classes + 1 empty
for (i in 1:5) {
  coef_vec <- as.matrix(coefs[[i]])[-1, ]  # exclude intercept
  non_zero <- coef_vec != 0
  if (sum(non_zero) > 0) {
    barplot(coef_vec[non_zero], 
            names.arg = rownames(coefs[[1]])[-1][non_zero],
            main = class_names[i],
            las = 2, 
            cex.names = 0.6,
            col = ifelse(coef_vec[non_zero] > 0, "darkgreen", "darkred"))
  }
}
par(mfrow = c(1, 1))  # reset
```


```{r}
# Calculate Test Accuracy
preds_min <- predict(cv_lasso, newx = x_test, s = "lambda.min", type = "class")
accuracy_min <- mean(as.character(preds_min) == as.character(y_test))
cat("Test accuracy (lambda.min):", accuracy_min, "\n")

# Calculate CM
preds_min <- factor(preds_min, levels = levels(y_test))
conf_mat_min <- confusionMatrix(preds_min, y_test, mode = "prec_recall")
print(conf_mat_min)

# Calculate Class Error
class_error_min <- 1 - diag(conf_mat_min$table) / colSums(conf_mat_min$table)
print(class_error_min)

# AUC Vectors
probs_min <- predict(cv_lasso, newx = x_test, s = "lambda.min", type = "response")
probs_min <- probs_min[,,1]
colnames(probs_min) <- levels(y_test)

# Repeat
preds_1se <- predict(cv_lasso, newx = x_test, s = "lambda.1se", type = "class")
accuracy_1se <- mean(as.character(preds_1se) == as.character(y_test))
cat("Test accuracy (lambda.1se):", accuracy_1se, "\n")

preds_1se <- factor(preds_1se, levels = levels(y_test))
conf_mat_1se <- confusionMatrix(preds_1se, y_test, mode = "prec_recall")
print(conf_mat_1se)

class_error_1se <- 1 - diag(conf_mat_1se$table) / colSums(conf_mat_1se$table)
print(class_error_1se)

probs_1se <- predict(cv_lasso, newx = x_test, s = "lambda.1se", type = "response")
probs_1se <- probs_1se[,,1]
colnames(probs_1se) <- levels(y_test)
```

```{r}
# Load model results (done for every model)
model_results <- rbind(model_results, data.frame(
  Model = "LASSO Min Lambda",
  Accuracy = accuracy_min,
  Precision = mean(conf_mat_min$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(conf_mat_min$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(conf_mat_min$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(y_test, probs_min)
))

model_results <- rbind(model_results, data.frame(
  Model = "LASSO 1se Lambda",
  Accuracy = accuracy_1se,
  Precision = mean(conf_mat_1se$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(conf_mat_1se$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(conf_mat_1se$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(y_test, probs_1se)
))
```

```{r}
# Create confusion matrix (done for every model)
cm_table <- as.data.frame(conf_mat_min$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "darkorange") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap (LASSO Min Lambda)",
       x = "Actual Class", y = "Predicted Class")
```
```{r}
cm_table <- as.data.frame(conf_mat_1se$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "skyblue") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap (LASSO Lambda 1se)",
       x = "Actual Class", y = "Predicted Class")
```



## PCA
```{r}
# Set PCA statistics
pca_result <- prcomp(df_scaled[, -ncol(df_scaled)], center = TRUE, scale. = TRUE)
explained_var <- summary(pca_result)$importance[2, ]
num_pcs <- which(cumsum(explained_var) >= 0.9)[1]

# Set PCA dataframe
df_pca <- data.frame(pca_result$x[, 1:num_pcs], response = df_scaled$response)
train_data_pca <- df_pca[train_index, ]
test_data_pca  <- df_pca[-train_index, ]

# PCA with CV
train_control <- trainControl(method = "cv", number = 10,
                              summaryFunction = multiClassSummary)

model_pca_cv <- train(response ~ ., data = train_data_pca,
                      method = "multinom",
                      trControl = train_control,
                      trace = FALSE)

print(model_pca_cv)
cat("Cross-validated Accuracy (training set):", max(model_pca_cv$results$Accuracy), "\n")

# Set our fit model
model_pca <- multinom(response ~ ., data = train_data_pca, trace = FALSE)

preds_pca <- predict(model_pca, newdata = test_data_pca)
preds_pca <- factor(preds_pca, levels = levels(test_data_pca$response))

# Calc CM
cm_pca <- confusionMatrix(preds_pca, test_data_pca$response, mode = "prec_recall")
print(cm_pca)

test_acc <- mean(as.character(preds_pca) == as.character(test_data_pca$response))
cat("Test Accuracy (PCA-based model):", test_acc, "\n")

loadings <- pca_result$rotation[, 1:num_pcs]
print("PCA Loadings (contribution of original variables):")
print(loadings)
```


```{r}
# Run it
train_control <- trainControl(method = "cv", number = 10,
                              summaryFunction = multiClassSummary)

model_pca_cv <- train(response ~ ., data = train_data_pca,
                      method = "multinom",
                      trControl = train_control,
                      trace = FALSE)

cat("Cross-validated accuracy (PCA-based model, training set):",
    max(model_pca_cv$results$Accuracy), "\n")

model_pca <- multinom(response ~ ., data = train_data_pca, trace = FALSE)

preds_pca <- predict(model_pca, newdata = test_data_pca)
preds_pca <- factor(preds_pca, levels = levels(test_data_pca$response))

cm_pca <- confusionMatrix(preds_pca, test_data_pca$response, mode = "prec_recall")
print(cm_pca)

test_acc_pca <- mean(as.character(preds_pca) == as.character(test_data_pca$response))
cat("Test Accuracy (PCA-based model):", test_acc_pca, "\n")

probs_pca <- predict(model_pca, newdata = test_data_pca, type = "probs")
probs_pca <- as.matrix(probs_pca)

colnames(probs_pca) <- colnames(probs_pca)
probs_pca <- probs_pca[, levels(test_data_pca$response), drop = FALSE]
```
```{r}
# Use first 2 PCs from your pca_result (even if you used more in model)
pca_scores_train <- data.frame(
  PC1 = pca_result$x[train_index, 1],
  PC2 = pca_result$x[train_index, 2],
  response = df_scaled$response[train_index],
  Dataset = "Train"
)

pca_scores_test <- data.frame(
  PC1 = pca_result$x[-train_index, 1],
  PC2 = pca_result$x[-train_index, 2],
  response = df_scaled$response[-train_index],
  Dataset = "Test"
)

pca_scores_all <- rbind(pca_scores_train, pca_scores_test)
var_pc1 <- round(explained_var[1] * 100, 1)
var_pc2 <- round(explained_var[2] * 100, 1)

ggplot(pca_scores_all, aes(x = PC1, y = PC2, color = response, shape = Dataset)) +
  geom_point(size = 2.5, alpha = 0.7) +
  stat_ellipse(data = subset(pca_scores_all, Dataset == "Train"), 
               aes(group = response), level = 0.95, alpha = 0.3) +
  theme_minimal() +
  labs(title = "PCA Score Plot - First Two Principal Components",
       x = paste0("PC1 (", var_pc1, "%)"),
       y = paste0("PC2 (", var_pc2, "%)")) +
  scale_shape_manual(values = c("Train" = 16, "Test" = 17))
```
```{r}
# Extract loadings for first 2 PCs
loadings_df <- as.data.frame(pca_result$rotation[, 1:2])
loadings_df$feature <- rownames(loadings_df)
loadings_df$importance <- sqrt(loadings_df$PC1^2 + loadings_df$PC2^2)

# Select top features to label (avoid overcrowding)
loadings_top <- loadings_df[order(loadings_df$importance, decreasing = TRUE)[1:min(15, nrow(loadings_df))], ]

ggplot(loadings_df, aes(x = PC1, y = PC2)) +
  geom_segment(aes(xend = 0, yend = 0), alpha = 0.3, color = "gray60") +
  geom_point(alpha = 0.3) +
  geom_text(data = loadings_top, 
            aes(label = feature, color = importance),
            size = 3, fontface = "bold") +
  scale_color_gradient(low = "darkblue", high = "darkred") +
  theme_minimal() +
  labs(title = "PCA Loadings - Feature Contributions to PC1 and PC2",
       subtitle = paste0("Top ", nrow(loadings_top), " features labeled by importance"),
       x = paste0("PC1 (", var_pc1, "%)"),
       y = paste0("PC2 (", var_pc2, "%)"),
       color = "Importance") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  theme(legend.position = "bottom")
```
```{r}
# Extract coefficients from your multinom model
coef_matrix <- coef(model_pca)

# Get PC coefficient names (they should be named PC1, PC2, etc.)
pc_cols <- grep("^PC", colnames(coef_matrix), value = TRUE)

# Extract only PC coefficients
pc_coefs <- coef_matrix[, pc_cols, drop = FALSE]

# Calculate mean absolute coefficient across all classes
pc_importance <- colMeans(abs(pc_coefs))

# Create dataframe for plotting
pc_importance_df <- data.frame(
  PC = names(pc_importance),
  Importance = pc_importance
)

# Extract PC numbers for proper ordering
pc_importance_df$PC_num <- as.numeric(gsub("PC", "", pc_importance_df$PC))
pc_importance_df <- pc_importance_df[order(pc_importance_df$PC_num), ]

# Plot
ggplot(pc_importance_df, aes(x = reorder(PC, PC_num), y = Importance)) +
  geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.8) +
  geom_text(aes(label = round(Importance, 3)), vjust = -0.5, size = 3) +
  theme_minimal() +
  labs(title = "Average Importance of PCs in Multinomial Model",
       subtitle = paste0("Based on mean absolute coefficients across all classes (", 
                         num_pcs, " PCs used)"),
       x = "Principal Component",
       y = "Mean |Coefficient| Across Classes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  expand_limits(y = max(pc_importance_df$Importance) * 1.15)
```

```{r}
cm_table <- as.data.frame(cm_pca$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "darkorange") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap (PCA)",
       x = "Actual Class", y = "Predicted Class")
```

```{r}
model_results <- rbind(model_results, data.frame(
  Model = "PCA",
  Accuracy = test_acc_pca,
  Precision = mean(cm_pca$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(cm_pca$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(cm_pca$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(test_data_pca$response, probs_pca)
))
```


## SVM Radial and Linear
```{r}
set.seed(67)
# CV SVM
svm_linear <- train(
  response ~ ., 
  data = train_data,
  method = "svmLinear2",
  trControl = trainControl(
    method = "cv",
    number = 10,
    classProbs = TRUE,
    summaryFunction = multiClassSummary
  ),
  preProcess = c("center", "scale","pca"),
  tuneGrid = expand.grid(cost = c(0.01, 0.1,1, 10)),
  probability = TRUE
)

print(svm_linear)
cat("Best cost (C):", svm_linear$bestTune$C, "\n")
cat("Cross-validated accuracy (training set):", max(svm_linear$results$Accuracy), "\n")

preds_svm_linear <- predict(svm_linear, newdata = test_data)
preds_svm_linear <- factor(preds_svm_linear, levels = levels(test_data$response))

cm_svm_linear <- confusionMatrix(preds_svm_linear, test_data$response, mode = "prec_recall")
print(cm_svm_linear)

test_acc_linear <- mean(as.character(preds_svm_linear) == as.character(test_data$response))
cat("Test Accuracy (SVM Linear):", test_acc_linear, "\n")

probs_svm_linear <- predict(svm_linear, newdata = test_data, type = "prob")
probs_svm_linear <- as.matrix(probs_svm_linear)

colnames(probs_svm_linear) <- colnames(probs_svm_linear)
probs_svm_linear <- probs_svm_linear[, levels(test_data_pca$response), drop = FALSE]

model_results <- rbind(model_results, data.frame(
  Model = "SVM Linear",
  Accuracy = test_acc_linear,
  Precision = mean(cm_svm_linear$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(cm_svm_linear$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(cm_svm_linear$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(test_data$response, probs_svm_linear)
))
```

```{r}
# Get variable importance
var_imp <- varImp(svm_linear, scale = FALSE)

# Extract importance based on its structure
imp_values <- var_imp$importance

# Extract feature names and scores properly
if(is.matrix(imp_values) || is.data.frame(imp_values)) {
  # If multiple columns (multiclass), take row means
  if(ncol(imp_values) > 1) {
    imp_scores <- rowMeans(abs(imp_values), na.rm = TRUE)
  } else {
    imp_scores <- imp_values[, 1]
  }
  feature_names <- rownames(imp_values)
} else {
  # If it's a vector
  imp_scores <- as.numeric(imp_values)
  feature_names <- names(imp_values)
}

# Create dataframe
imp_df <- data.frame(
  Feature = feature_names,
  Importance = imp_scores,
  row.names = NULL
)




# Sort by importance
imp_df <- imp_df[order(imp_df$Importance, decreasing = TRUE), ]

# Check if features are PC names or original features
if(grepl("^PC", imp_df$Feature[1])) {
  cat("Features are PC names - back-projecting to original features...\n")
  
  # Extract PC importance
  n_pcs <- nrow(imp_df)
  pc_imp <- imp_df$Importance[1:n_pcs]
  
  # Get PCA rotation
  pca_rot <- svm_linear$preProcess$rotation
  
  # Calculate original feature importance
  if(ncol(pca_rot) >= n_pcs) {
    pca_rot_used <- pca_rot[, 1:n_pcs, drop = FALSE]
    feature_imp <- as.vector(abs(pca_rot_used) %*% matrix(pc_imp, ncol = 1))
    feature_names <- rownames(pca_rot_used)
    
    imp_df <- data.frame(
      Feature = feature_names,
      Importance = feature_imp
    )
    imp_df <- imp_df[order(imp_df$Importance, decreasing = TRUE), ]
  }
}

# Normalize to 0-100 scale for better interpretation
imp_df$Importance_norm <- 100 * imp_df$Importance / max(imp_df$Importance)

# Top 20 features
top_n <- min(20, nrow(imp_df))
imp_top20 <- imp_df[1:top_n, ]
imp_top20$Feature <- factor(imp_top20$Feature, 
                             levels = rev(imp_top20$Feature))

# Plot
ggplot(imp_top20, aes(x = Feature, y = Importance_norm)) +
  geom_segment(aes(xend = Feature, yend = 0), 
               color = "skyblue", linewidth = 1.5) +
  geom_point(size = 3, color = "darkblue") +
  geom_text(aes(label = round(Importance, 3)), hjust = -0.2, size = 3) +
  coord_flip() +
  theme_minimal() +
  labs(title = "SVM Linear of Combined Class - Top 20 Most Important Features",
       subtitle = paste0("Cost C = ", svm_linear$bestTune$cost,
                        " | Test Accuracy = ", round(test_acc_linear, 4)),
       x = "",
       y = "Relative Importance (0-100 scale)") +
  expand_limits(y = max(imp_top20$Importance_norm) * 1.15)
```







```{r}
# Repeat
svm_radial <- train(
  response ~ ., 
  data = train_data,
  method = "svmRadial",
  trControl = trainControl(
    method = "cv",
    number = 10,
    classProbs = TRUE,     
    summaryFunction = multiClassSummary
  ),
  tuneGrid = expand.grid(
    C = c(0.1, 1, 10),
    sigma = c(0.01, 0.05, 0.1)
  ),
  prob.model = TRUE          
)


print(svm_radial)
cat("Best parameters: C =", svm_radial$bestTune$C,
    " sigma =", svm_radial$bestTune$sigma, "\n")
cat("Cross-validated accuracy (training set):", max(svm_radial$results$Accuracy), "\n")


preds_svm_radial <- predict(svm_radial, newdata = test_data)
preds_svm_radial <- factor(preds_svm_radial, levels = levels(test_data$response))

cm_svm_radial <- confusionMatrix(preds_svm_radial, test_data$response, mode = "prec_recall")
print(cm_svm_radial)

test_acc_radial <- mean(as.character(preds_svm_radial) == as.character(test_data$response))
cat("Test Accuracy (SVM Radial):", test_acc_radial, "\n")

probs_svm_radial <- predict(svm_radial, newdata = test_data, type = "prob")
probs_svm_radial <- as.matrix(probs_svm_radial)

colnames(probs_svm_radial) <- colnames(probs_svm_radial)
probs_svm_radial <- probs_svm_radial[, levels(test_data$response), drop = FALSE]

model_results <- rbind(model_results, data.frame(
  Model = "SVM Radial Kernel",
  Accuracy = test_acc_radial,
  Precision = mean(cm_svm_radial$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(cm_svm_radial$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(cm_svm_radial$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(test_data$response, probs_svm_radial)
))
```

```{r}
cm_table <- as.data.frame(cm_svm_linear$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "#613fbaff") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap for Combined Class (SVM Linear)",
       x = "Actual Class", y = "Predicted Class")

cm_table <- as.data.frame(cm_svm_radial$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "skyblue") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap (SVM Radial)",
       x = "Actual Class", y = "Predicted Class")
```


## Random Forest
```{r}
set.seed(6767)
# CV for random forest
train_control <- trainControl(method = "cv", number = 10)

tune_grid <- expand.grid(mtry = c(2, 5, 7, 10, 15, 20))

results <- list()
ntree_values <- c(100, 250, 500)

for (nt in ntree_values) {
  rf_model <- train(response ~ ., data = train_data,
                    method = "rf",
                    trControl = train_control,
                    tuneGrid = tune_grid,
                    ntree = nt)
  
  results[[paste0("ntree_", nt)]] <- rf_model
  
  preds <- predict(rf_model, newdata = test_data)
  cm <- confusionMatrix(preds, test_data$response, mode = "prec_recall")
  
  # Print results
  cat("\nResults for ntree =", nt, ":\n")
  print(rf_model$results)
  cat("Best mtry:", rf_model$bestTune$mtry, "\n")
  cat("CV Accuracy:", max(rf_model$results$Accuracy), "\n")
}
```
```{r}
# Use best model as fit
rf_model <- train(response ~ ., data = train_data,
                  method = "rf",
                  trControl = trainControl(method = "cv", number = 10,
                                           summaryFunction = multiClassSummary),
                  tuneGrid = data.frame(mtry = results[["ntree_500"]]$bestTune$mtry),
                  ntree = 500)
preds_rf <- predict(rf_model, newdata = test_data)
cm_rf <- confusionMatrix(preds_rf, test_data$response, mode = "prec_recall")

probs_rf <- predict(rf_model, newdata = test_data, type = "prob")
probs_rf <- as.matrix(probs_rf)

colnames(probs_rf) <- colnames(probs_rf)
probs_rf <- probs_rf[, levels(test_data$response), drop = FALSE]

model_results <- rbind(model_results, data.frame(
  Model = "Base Random Forest",
  Accuracy = mean(preds_rf == test_data$response),   
  Precision = mean(cm_rf$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(cm_rf$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(cm_rf$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(test_data$response, probs_rf)
))

print(rf_model)
print(cm_rf)

# Print Variable Importance Plot
varImpPlot(rf_model$finalModel, n.var = 10, main = "Variable Importance for Best Random Forest Model")
```
```{r}
cm_table <- as.data.frame(cm_rf$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "skyblue") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap for Random Forest of Combined Class",
       x = "Actual Class", y = "Predicted Class")
```

## Upsampling
```{r}
set.seed(67)

# Cross validation for upsampling
train_control <- trainControl(method = "cv", number = 10,
                              sampling = "up",
                              summaryFunction = multiClassSummary)

rf_model_balanced <- train(response ~ ., data = train_data,
                           method = "rf",
                           trControl = train_control,
                           tuneGrid = data.frame(mtry = c(2, 5, 10, 12, 15,20)),
                           ntree = 500)

print(rf_model_balanced)
cat("Best mtry:", rf_model_balanced$bestTune$mtry, "\n")
cat("Cross-validated accuracy:", max(rf_model_balanced$results$Accuracy), "\n")
preds_rf_up <- predict(rf_model_balanced, newdata = test_data)
cm_rf_up <- confusionMatrix(preds_rf_up, test_data$response, mode = "prec_recall")

probs_rf_up <- predict(rf_model_balanced, newdata = test_data, type = "prob")
probs_rf_up <- as.matrix(probs_rf_up)

colnames(probs_rf_up) <- colnames(probs_rf_up)
probs_rf_up <- probs_rf_up[, levels(test_data$response), drop = FALSE]

model_results <- rbind(model_results, data.frame(
  Model = "Upsampling Random Forest",
  Accuracy = mean(preds_rf_up == test_data$response),  
  Precision = mean(cm_rf_up$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(cm_rf_up$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(cm_rf_up$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(test_data$response, probs_rf_up)
))
```

```{r}
cm_table <- as.data.frame(cm_rf_up$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "darkorange") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap Upsampling Random Forest for Combined Class",
       x = "Actual Class", y = "Predicted Class")
```




## Class Weights
```{r}
set.seed(12)
# Set class weights
class_counts <- table(train_data$response)
class_weights <- 1 / class_counts
class_weights <- class_weights / sum(class_weights)

# CV for class weights
rf_weighted <- randomForest(response ~ ., data = train_data,
                            ntree = 500,
                            mtry = 7,
                            importance = TRUE,
                            classwt = class_weights)

preds_rf_weighted <- predict(rf_weighted, newdata = test_data)

cm_rf_weighted <- confusionMatrix(preds_rf_weighted, test_data$response, mode = "prec_recall")
accuracy_rf_weighted <- mean(as.character(preds_rf_weighted) == as.character(test_data$response))
cat("Test Accuracy:", accuracy_rf_weighted, "\n")
print(cm_rf_weighted)

probs_rf_weighted <- predict(rf_weighted, newdata = test_data, type = "prob")
probs_rf_weighted <- as.matrix(probs_rf_weighted)

colnames(probs_rf_weighted) <- colnames(probs_rf_weighted)
probs_rf_weighted <- probs_rf_weighted[, levels(test_data$response), drop = FALSE]

model_results <- rbind(model_results, data.frame(
  Model = "Weighted Random Forest Non-CV",
  Accuracy = accuracy_rf_weighted,  
  Precision = mean(cm_rf_weighted$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(cm_rf_weighted$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(cm_rf_weighted$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(test_data$response, probs_rf_weighted)
))
```
```{r}
cm_table <- as.data.frame(cm_rf_weighted$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "forestgreen") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap Weighted Random Forest For Combined Class",
       x = "Actual Class", y = "Predicted Class")
```
```{r}
set.seed(12)

# Set class weights
class_counts <- table(train_data$response)
class_weights <- 1 / class_counts
class_weights <- class_weights / sum(class_weights)

# Calculate observation-level weights
obs_weights <- class_weights[as.character(train_data$response)]

# 10-fold CV with caret - Fixed version
rf_weighted_cv <- train(
  response ~ ., 
  data = train_data,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10,
    classProbs = TRUE,
    summaryFunction = multiClassSummary,
    verboseIter = FALSE
  ),
  ntree = 500,
  # Use tuneGrid instead of mtry parameter
  tuneGrid = data.frame(mtry = 7),  # Fixed mtry value
  importance = TRUE,
  weights = obs_weights
)

# CV results
cat("\n===== 10-Fold Cross-Validation Results =====\n")
print(rf_weighted_cv)
cat("Cross-validated Accuracy:", max(rf_weighted_cv$results$Accuracy), "\n")

# Test set predictions
preds_rf_weighted <- predict(rf_weighted_cv, newdata = test_data)

cm_rf_weighted <- confusionMatrix(preds_rf_weighted, test_data$response, mode = "prec_recall")
accuracy_rf_weighted <- mean(as.character(preds_rf_weighted) == as.character(test_data$response))

cat("\n===== Test Set Results =====\n")
cat("Test Accuracy:", accuracy_rf_weighted, "\n")
print(cm_rf_weighted)

probs_rf_weighted <- predict(rf_weighted_cv, newdata = test_data, type = "prob")
probs_rf_weighted <- as.matrix(probs_rf_weighted)

colnames(probs_rf_weighted) <- colnames(probs_rf_weighted)
probs_rf_weighted <- probs_rf_weighted[, levels(test_data$response), drop = FALSE]

model_results <- rbind(model_results, data.frame(
  Model = "Weighted Random Forest CV",
  Accuracy = accuracy_rf_weighted,  
  Precision = mean(cm_rf_weighted$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(cm_rf_weighted$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(cm_rf_weighted$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(test_data$response, probs_rf_weighted)
))
```

```{r}
# Original best model, so we check the VarImpPlot
varImpPlot(rf_weighted, main="RF Weighted Variable Importance")
```
```{r}
# Original best model, so we print a pretty tree
single_tree <- randomForest::getTree(rf_weighted, k = 1, labelVar = TRUE)

to_rpart <- function(tree, data) {
  response <- all.vars(formula(rf_weighted))[1]
  form <- as.formula(paste(response, "~ ."))
  rpart(form, data = data, control = rpart.control(cp = 0.0001))
}

rpart_tree <- to_rpart(single_tree, train_data)
rpart.plot(
  rpart_tree,
  type = 4,
  extra = 101,
  fallen.leaves = TRUE,
  box.palette = "RdYlGn",
  branch.lty = 3,
  shadow.col = "gray",
  main = "Decision Tree Extracted from Random Forest"
)
```



## XGBoost
```{r}
set.seed(9999)

# Cross Validation for XGBoost
train_control <- trainControl(method = "cv", number = 10)
tune_grid <- expand.grid(
  nrounds = c(100, 250, 500),
  max_depth = c(3, 5, 7),
  eta = c(0.01, 0.1, 0.3),
  gamma = 0,
  colsample_bytree = 0.8,
  min_child_weight = 1,
  subsample = 0.8
)


xgb_model <- train(
  response ~ ., data = train_data,
  method = "xgbTree",
  trControl = train_control,
  tuneGrid = tune_grid,
  verbose = FALSE   
)

print(xgb_model)
cat("Best parameters:\n")
print(xgb_model$bestTune)
cat("Cross-validated accuracy:", max(xgb_model$results$Accuracy), "\n")

preds_xgb <- predict(xgb_model, newdata = test_data)
preds_xgb <- factor(preds_xgb, levels = levels(test_data$response))

cm_xgb <- confusionMatrix(preds_xgb, test_data$response)
print(cm_xgb)
```

```{r}
set.seed(9999)
# Another cross validation
trainControl(method = "cv", number = 10, summaryFunction = multiClassSummary)


tune_grid <- expand.grid(
  nrounds = 100,
  max_depth = 5,
  eta = 0.01,
  gamma = 0,
  colsample_bytree = 0.8,
  min_child_weight = 1,
  subsample = 0.8
)

xgb_model <- train(response ~ ., data = train_data,
                   method = "xgbTree",
                   trControl = trainControl(method = "cv", number = 10),
                   tuneGrid = tune_grid,
                   verbose = FALSE)


cat("Cross-validated accuracy:", max(xgb_model$results$Accuracy), "\n")

preds_xgb <- predict(xgb_model, newdata = test_data)
preds_xgb <- factor(preds_xgb, levels = levels(test_data$response))

cm_xgb <- confusionMatrix(preds_xgb, test_data$response, mode = "prec_recall")
print(cm_xgb)

probs_xgb <- predict(xgb_model, newdata = test_data, type = "prob")
probs_xgb <- as.matrix(probs_xgb)

colnames(probs_xgb) <- colnames(probs_xgb)
probs_xgb <- probs_xgb[, levels(test_data$response), drop = FALSE]

model_results <- rbind(model_results, data.frame(
  Model = "XGBoost",
  Accuracy = max(xgb_model$results$Accuracy),
  Precision = mean(cm_xgb$byClass[,"Precision"], na.rm = TRUE),
  Recall = mean(cm_xgb$byClass[,"Recall"], na.rm = TRUE),
  F1 = mean(cm_xgb$byClass[,"F1"], na.rm = TRUE),
  AUC = macro_auc(test_data$response, probs_xgb)
))
```

```{r}
cm_table <- as.data.frame(cm_xgb$table)
ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "darkorange") +
  theme_minimal() +
  labs(title = "Confusion Matrix Heatmap (XGBoost)",
       x = "Actual Class", y = "Predicted Class")
```

## Model Results
```{r}
print(model_results)
model_results_combined <- model_results
```





